Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-25 src/1.Introduction.tex
Small edits

Teaching generic parallel programming today is difficult. Educators strive to balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do programming models offer enough abstractions to support modern hardware architectures sufficiently . Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine specification. We would like to compare this to the support of Von Neumann architectures in programming languages today. We define an abstract machine model later in this work. The lack of such parallel programming models motivated us to develop Kokkos.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively between national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}. We list its key properties as follows.

@@

The objective of this paper is to present Kokkos from two perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. These insights may be  useful in parallel programming classes that introduce programming models and reason about design choices.

@@

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

>>>>

Teaching generic parallel programming today is difficult. Educators strive to balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do many programming models offer enough abstractions to support modern hardware architectures sufficiently. Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine specification. We would like to compare this to the support of Von Neumann architectures in programming languages today.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively by a number of national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namelyexecution patterns and data structures which it maps to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}. We list its key properties as follows.

@@

The objective of this paper is to present Kokkos from three perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. These insights may be  useful in parallel programming classes that introduce programming models and reason about design choices.

@@

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions. We show a set of interfaces that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

### 2020-01-25 src/2.GenericProgramming.tex
Small edits

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often achieved using \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming syntaxes. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantic of describing a parallel loop construct. It is up to the programming model specification to make further definitions regarding separation of concern.

>>>>

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show two code examples using the aforementioned parallel programming syntaxes. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantic of describing a parallel loop construct. It is up to the programming model specification to make further definitions regarding separation of concern.
